---
title: "Lec 2 R Functions And The Grammar Of Visualization"
author: "Xinan Wang"
date: "2023-01-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functions

r functions are called like this: function_name(arg1 = val1, arg2 = val2, ...)

```{r}

sum(1,2)

```

## Create Functions

variable <- function(arg1, arg2, ...) {
  your expression / algorithm
}

```{r}

hello <- function() {
  print("Hello World!")
}

hello()

```

## Name Masking

```{r}

f <- function() {
  x <- 1
  y <- 2
  c(x, y)
}
f()

rm(f)

```

If a name isn't defined inside a function, R will look one level up

```{r}

x <- 2
g <- function() {
  y <- 1
  c(x, y)
}
g()

rm(g)

```

## Functions v.s. Variables

```{r}

n <- function(x)
  x / 2
o <- function() {
  n <- 10
  n(n)
}
o()

rm(n, o)

```

```{r}

f <- function(x) {
  f <- function(x) {
    f<- function(x) {
      x ^ 2
    }
    f(x) + 1
  }
  f(x) * 2
}
f(10)
```

## Every operation is a function call

```{r}

x <- 1
y <- 2
x+y

`+`(x,y)

```

```{r}

for (i in 1:2) print(i)

`for`(i, 1:2, print(i))

```

## Anonymous Functions

function(x)3()

(function(x)3)()

## Return Values

```{r}

double.num <- function(x) {
  x * 3
  print("hello")
  x * 2
}
double.num(5)

double.num <- function(x) {
  return(x * 3)
  print("hello")
  return(3)
}
double.num(5)

```

## Functions with if and else

```{r}

if.one <- function(x) {
  if(x == 1) {
    print("True")
  }
  else {
    print("False")
  }
}
if.one(1)
if.one(2)


if.one <- function(x){
  ifelse(x==1, "TRUE", "FALSE")
}
if.one(1)
if.one(2)

```

## Switch

```{r}

multipleCases <- function(x) {
  switch(x,
         a = "first",
         b = "second",
         z = "last",
         c = "third",
         d = "other")
}
multipleCases("a")

```

## ...

```{r}

f <- function(a,b,c) {
  data.frame(a,b,c)
}
f(a = 1, b = 2, c = 3)

```

```{r}

f <- function(...) {
  data.frame(...)
}
f(a = 1, b = 2, c = 3, d = 5, e = 7)

```


# For Loops

```{r}

for (i in 1:3) {
  print(i)
}

```

```{r}

fruit <- c("apple", "banana", "pomegranate")

fruitlength <- rep(NA, length(fruit))
fruitlength

names(fruitlength) <- fruit
fruitlength

```

```{r}

for (i in fruit) {
  fruitlength[i] <- nchar(i)
}
fruitlength

```

# Apply

## apply

Must be used on a matrix, meaning all the elements must be of the same type whether they are character, numeric or logical.

```{r}

theMatrix <- matrix(1:9, nrow = 3)
theMatrix
# sum the row
apply(theMatrix, 1, sum)

apply(theMatrix, 2, sum)

rowSums(theMatrix)

colSums(theMatrix)

```

## lapply, sapply

Basic grammar

lapply(x, FUN, ...)
sapply(x, FUN, ...)

lapply works by applying a function to each element of a list and returning the results as a list

```{r}

theList <- list(A = 1:3, B = 1:5, C = -1:1, D = 2)
lapply(theList, sum)

```

sapply is a user-friendly version and wrapper of lapply by default returning a vector

```{r}

theList <- list(A = 1:3, B = 1:5, C = -1:1, D = 2)
sapply(theList, sum)

```

## mapply

Mapply: applies a function to each elements of multiple lists

```{r}

firstList <-
  list(A = matrix(1:16, 4),
       B = matrix(1:16, 2),
       C = data.frame(1:5))
secondList <-
  list(A = matrix(1:16, 4),
       B = matrix(1:16, 8),
       C = data.frame(15:1))
mapply(identical, firstList, secondList)


simpleFunc <- function(x, y) {
  nrow(x) + nrow(y)
}
mapply(simpleFunc, firstList, secondList)

```

# Graphics

## Basic Plots

• Bar chart (categorical variable vs continuous variable)

• Scatter plot (continuous variable vs continuous variable)

• Line chart (continuous variable vs continuous variable)

• Box plot (categorical variable vs continuous variable)

• Histogram (continuous variable vs continuous variable)

Title of graph, scale, coordinate system, context, visual cues

Components of the plots:
• Layers: Dataset;
          Aesthetic mapping(color, shape, size, etc)
          Statistical transformation
          Geometric object(line, bar, dots, etc)
          Position adjustment
• Scale(optional)
• Coordinate system
• Faceting(optional)
• Defaults

## ggplot

ggplot(data = <DATASET>,

      mapping = aes(<Mappings>) + 
      
        layer(geom = <GEOM>,
        
              stat = <STAT>,
              
              position = <POSITION>) + 
              
        <SCALE_FUNCTION>() + 
        
        <COORDINATE_FUNCTION>() +
        
        <FACET_FUNCTION>()

```{r}

df <- data.frame(
  A = c(2, 1, 4, 9),
  B = c(3, 2, 5, 10),
  C = c(4, 1, 15, 80),
  D = c("a", "a", "b", "b")
)
df
```

```{r}

library(ggplot2)

df <- data.frame(
  A = c(2, 1, 4, 9),
  B = c(3, 2, 5, 10),
  C = c(4, 1, 15, 80),
  D = c("a", "a", "b", "b")
)

ggplot(data = df,
       mapping = aes(x = A, y = C, shape = D)) + 
  layer(geom = "point",
        stat = "identity",
        position = "identity") +
  scale_x_continuous() +
  scale_y_continuous() +
  coord_cartesian() + 
  facet_null()
 
```

```{r}

ggplot(data = df, mapping = aes(x = A, y = C, shape = D)) + 
  geom_point() +
  facet_grid( ~ D)

```

Dataset : Fuel economy in cars

```{r}

library(tidyverse)

head(mpg, 5)

# Interviewing data

mpg[,c("displ","hwy")]

# Creating base

ggplot(data = mpg)

# Creating plot

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))

```

### Aesthetic Mappings

Aesthetic mappings: We can change levels of size, shape, color, fill, alpha etc. inside of aes()

```{r}

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class))

```

Aesthetic mappings: outside of aes()

• The name of a color as character string

• The size of a point in mm

• The shape of a point as a number

### geom_() functions

• geom_bar: bar chart

• geom_histogram: histogram

• geom_point: scatterplot

• geom_qq: quantile-quantile plot

• geom_boxplot: boxplot

• geom_line: line chart

```{r}

geom_line

```

• Use geom_ function to make variables visible on the screen

• Use stat_ function and define geom shape as an argument inside geom_.

• Or use geom_ and define statistical transformation as an argument inside stat_.

Common geom with position adjustments:

geom_point, geom_jitter

### Scale syntax

labs() xlab() ylab() ggtitle() lim() xlim() ylim()
scale_colour_brewer() scale_colour_continuous()

```{r}

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = hwy)) + 
  scale_colour_continuous(type = "viridis")

```

```{r}

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  coord_polar()

```

### Faceting

grid:

```{r}

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(.~year)

```

wrap:

```{r}

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(.~cyl)

```

## A more complicated embedded grammer

```{r}

troops <- read.table("minard-troops.txt", header = T)
cities <- read.table("minard-cities.txt", header = T)

head(troops)

head(cities)

```

```{r}

plot_troops <- ggplot(troops, aes(long, lat)) +
  geom_path(aes(size = survivors, colour = direction, group = group))
plot_troops

```

```{r}

plot_both <- plot_troops +
  geom_text(aes(label = city), size = 4, data = cities)
plot_both

```

```{r}

library(mapproj)

plot_polished <- plot_both +
  scale_size(
    range = c(0, 12),
    # breaks = c(10000, 20000, 30000),
    # labels = c("10,000", "20,000", "30,000")
  ) +
  scale_color_manual(values = c("tan", "grey50")) + 
  coord_map() +
  labs(title = "Map of Napoleon's Russian campaign of 1812") + 
  theme_void() +
  theme(legend.position = "none")
plot_polished

```