---
title: "Lec 1(1) Basic of R I"
author: "Xinan Wang"
date: "2023-01-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Install Required Packages

install.packages('installr')

install.packages('tidyverse')

library(tidyverse)

install.packages('rmarkdown')


if (!requireNamespace('devtools'))

  install.packages('devtools')
  
devtools::install_github('rstudio/rmarkdown')


install.packages('tinytex')

tinytex::install_tinytex()


install.packages('ggplot2')

install.packages('nycflights13')

install.packages('gapminder')


libs <- c('ggplot2','nycflights13','gapminder')


x <- sapply(libs, function(x)

  if (!require(
    x,
    character.only = T,
    warn.conflicts = F,
    quietly = T
  ))
    install.packages(x))


# Basic Math

```{r}

1 + 1
1/(1+1)

```

# Function in R

```{r}

if (sample(10,1)%in%c(1:9)) {
  print('My students love me')
} else {
  'They hate me'
}
  
```

# Variable

```{r}

x <- 13
x

x = 13
x

assign('x',13)
x

# Remove variables
x <- 13
rm(x)

```

# Data Types

```{r}

x <- 13
class(x)

x <- 13
is.numeric(x)

x <- 3
is.integer(x)

x <- 3L
is.integer(x)

is.numeric(x)
```

# Character

R has 2 primary ways of handling character data: character and factor

```{r}

# Store 'RStudio' to x as character data
x <- 'RStudio'
x

# Store 'RStudio' to x as factor data by using factor function
x <- factor('RStudio')
x <- factor(c('a','c','D'))
x <- factor(c(1,2,3))
x

# Length
x <- 'RStudio'
nchar(x)  ## nchar can only be used for character and numeric data

y <- 365
nchar(y)

```

# Date

```{r}

date <- as.Date('2022-01-17')
date

class(date)

as.numeric(date)

as.numeric(2022-01-17)

```

# Data Wrangling

install.packages('stringr') # For string
install.packages('lubridate') # For date
install.packages('forcats') # For factor

## Logical

```{r}

x <- 13
is.numeric(x)

x1 <- 'I am awesome'
is.numeric(x1)

is.character(x1)

is.na(x1)

x <- TRUE
class(x)

is.logical(x)

TRUE * 5

c(TRUE, TRUE, TRUE)*1

FALSE * 5

# Comparison

1 == 1

1 != 1

1 < 1

1 <= 2

'i' == 'IE6600'

'i' < 'IE6600'

```

## Vector

A vector is a collection of elements, all of the same type. A vector cannot be of mixed type.

Vectors don't have dimensions

Column or row vectors can be represented as one-dimensional matrices.

```{r}

x <- c('IE6600', 'Data', 'Visulization')

class(x)

y <- c(2,3)

class(y)

x <- c(1,2,3,4,5)
x

y <- c(1:5)
y

x * 3

y^2

sqrt(x)

length(x)

length(y)

x + y

x * y

x / y

x^y

x <- c(1:5)
x + c(1,2) ## 1+1 2+2 3+1 4+2 5+1

x <= 3

nchar(x)

c('I', 'am', 'beautiful')

c(One='I', Two='am', Three='beautiful')

x <- c(5:10)

x[2]

x[c(1,3)]

x[1:3]

```


## Factor

```{r}

x <- c('I','am','awesome')
x

x <- as.factor(x)
x

x <- factor(x, levels=c('I','awesome','am'))
x

```

## Missing Data

R has 2 types of missing data: NA and NULL. Similar but behave differently

### NA

NA will often be seen as just another element of a vector. is.na tests each element of a vector for missingness

```{r}

x <- c(1,2,3,NA,5)
x

length(x)

is.na(x)

```

### NULL

NULL is nothingness. Functions can sometimes return NULL and their arguments can be NULL

NULL is often returned by expressions and functions whose value is undefined

NULL is atomical and cannot exist within a vector.

```{r}

x <- c(1,2,3,NULL,5)

length(x)

is.null(x)

```

# Call Functions

```{r}

mean(c(1,2,3))

sum(c(1,2,3))

```

# Pipe

```{r}

library(magrittr)
library(dplyr)

y <- c(1,2,3)
mean(y)

c(1,2,3) %>% mean

```

# Advanced Data Structures

The most common are data frame, matrix, and list.

## Data Frames

R organized data.frame, each column is actually a vector, each of which has the same length. Within a column each element must be of the same type

```{r}

x <- 1:5
y <- -1:3
a <- c('I','am','an','awesome','student')

xya.df <- data.frame(x,y,a)
xya.df

xya.df <- data.frame(1:5, -1:3, a)
xya.df

xya.df <- data.frame(first = x, second = y, third = a)
xya.df

names(xya.df) <- c(x='first', y='second', a='third')
xya.df

nrow(xya.df)

ncol(xya.df)

dim(xya.df)

names(xya.df)

head(xya.df)

head(xya.df, n=3)

xya.df$second

xya.df$newColumn <- c(1:5)
xya.df

xya.df[5,2] # 5th row, 2th col

xya.df[5,] # 5th row

xya.df[c(1,2),3] # 1st & 2nd row, 3rd col

class(xya.df[, 'third'])

xya.df[, 'third', drop=FALSE]
class(xya.df[, 'third', drop=FALSE])

xya.df[xya.df$second > 1, ]

xya.df[xya.df$second > 1, 2:3]

```

## List

A list can contain all numeric or characters or a mix of the 2 or data.frames or recursively other lists.

Lists are created with the list function where each argument to the function becomes an element of the list

```{r}

list(1,2,3) # Create a 3 element list

list(c(1,2,3)) # Create a single element list

list(c(1,2,3), 1:4) # Create a 2 element list

b <- c('master','phD','undergraduate','others','master')
list(c(1:3),b)

d <- list(c(1:3),b)

names(d) <- c('number','degree')
d

list1 <- list(number=1:3, degree=b)
list1

list1 <- list(number=1:3,degree=b)

list1[[1]]

list1[1]

d <- list(number = 1:3, degree = b)
length(d)

d[[3]] <- c(7,7,7,7)
d[['student']] <- c('John','Peter','Tome','Jerry')
d

```

## Matrix

```{r}

A <- matrix(1:6, nrow = 3)
A

nrow(A)

ncol(A)

B <- matrix(2:7, nrow = 3)
B

nrow(B)

ncol(B)

A == B

A * B

```

# Data Input to R

```{r}

readurl <- 'http://www.jaredlander.com/data/Tomato%20First.csv'
tomato <- read.csv(readurl)

tomato

```

## CSV data

library(tidyverse)

library(readr)


tomato <- read_csv('tomato.csv')

## Excel data

install.packages('readxl')

library(readxl)


tomato <- read_excel('excel location')

## Other types of data

read.spss => SPSS

read.dta => Stata

read.ssd => SAS

read.octave => Octave

read.mtp => Minitab

read.systat => Systat